### 2019-02-26 summary/paper-summary.tex
added the first lines of the paper summary and a sample set of the results

\documentclass{article}

\usepackage{hyperref}
\usepackage{xcolor}

\begin{document}

	\title{Leveraged Volume Sampling - Summary}
	\author{Ohad Ohayon}
	\pagenumbering{gobble}
	\maketitle
	\newpage

	\pagenumbering{arabic}

    \tableofcontents
    \newpage

    \section{Overview}
        This document summarises the contribution of the {\color{blue}\href{https://arxiv.org/abs/1802.06749}{\underline{paper}}} \textit{"Leveraged volume sampling for linear regression."}.
        It follows the structure of the paper itself, summarizing each of it's main sections.
        Finally, I will demonstrate my attempt at reproducing the results in the paper.

    \section{Introduction}
        \subsection{The Problem}
            Assume a linear regression problem where the feature vectors are the rows of an $n$ by $d$ full rank matrix $X=(x_{i}^T)$,
            the labels are the column vector $y$ and the optimal solution is $w^\ast=X^+y$.
            Define the average loss function of a solution $w$ to be $L(w)=frac{1}_{n}||Xw-y||_{2}^2$.
            In many real-world scenarios acquiring labels for some linear regression problems can be an expensive operation,
            although the regression features are readily available (as they may be just the parameters for an experiment).
            You may be accessible to the labels but you must limit the amount you request in order to deliver results within
            reasonable resource constraints.
            Assume the rows of such a sub-sampled problem to be $S$, the sub-sampled matrix and labels vector to be $X_{S}$
            and $y_{S}$ respectively and the sub-sampled solution to be $w_{S}=X_{S}^+y_{S}$.
            We'd like an effective algorithm to choose $S$, based on the features alone, whose corresponding solution $w_{S}$
            will hold the following properties with regard to the optimal solution $w^\ast$:
            \begin{itemize}
                \item $w_{S}$ is unbiased, i.e.: $E[w_{S}]=w^\ast$.
                \item The error for $w_{S}$ can be bounded by $L(w_{S}) < (1+\epsilon)L(w^\ast)$ for an $\epsilon$ of choice.
                \item Can be executed efficiently.
            \end{itemize}

\end{document}

### 2019-02-26 summary/paper-summary.tex
added a summary of leverage score sampling to the introduction of the paper summary

@@
 \usepackage{xcolor}
+\usepackage{amsmath}
 
@@
             the labels are the column vector $y$ and the optimal solution is $w^\ast=X^+y$.
-            Define the average loss function of a solution $w$ to be $L(w)=frac{1}_{n}||Xw-y||_{2}^2$.
+            Define the average loss function of a solution $w$ to be $L(w)=\frac{1}{n}||Xw-y||_{2}^2$.
             In many real-world scenarios acquiring labels for some linear regression problems can be an expensive operation,
@@
 
+        \subsection{Leverage Score Sampling}
+            The leverage score $l_{i}$ of row $i$ of matrix $X=(x_{i}^T)$ is given by $l_{i}=x_{i}^T(X^TX)^{-1}x_{i}$ and those
+            values sum up to $d$.
+            \textit{Leverage Score Sampling} suggests selecting the multiset $S$ of $k$ rows i.i.d from $X$ with probability:
+
+            \begin{equation}
+            P(S\in{[n]^k})=\prod_{i=1}^{k}\frac{l_{i}}{d}
+            \end{equation}
+
+            With this procedure we are guaranteed\footnote{\label{leverage_score_performance_1}"Reverse iterative volume
+            sampling for linear regression" P.21} that it is sufficient to choose
+            $O(d\log{d}+\frac{d}{\epsilon})$ to obtain $L(w_{S}) < (1+\epsilon)L(w^\ast)$ with high probability.
+            All in all this is a pretty efficient and fast algorithm.
+            However, this multiplicative bound doesn't hold in expectation and the approximate solution $w_{S}$ is biased (a
+            trait of i.i.d sampling algorithms). This means that you can't control for the error by repeating the process
+            and averaging the estimators.
+
 \end{document}

### 2019-02-26 summary/paper-summary.tex
added an introduction summary of volume sampling

@@
             and $y_{S}$ respectively and the sub-sampled solution to be $w_{S}=X_{S}^+y_{S}$.
+
             We'd like an effective algorithm to choose $S$, based on the features alone, whose corresponding solution $w_{S}$
@@
             \end{itemize}
+
+            For the rest of this summary we'll assume $X$ is in general position unless otherwise specified.
 
@@
 
+        \subsection{Volume Sampling}
+            Volume sampling is currently the main alternative to Leverage Score Sampling.
+            It is a non-i.i.d sampling method that attempts to sample a "diverse" subset of the rows of $X$,
+            where "diversity" is in the sense of trying to maintain as much of the volume spanned by the
+            columns of $X$.
+            \textit{Volume Sampling} suggest selecting the set $S$ of $k$ distinct rows from $X$ with the following probability:
+
+            \begin{equation}
+                P(S\in{\binom{[n]}{k}})=\frac{det(X_{S}^TX_{S})}{\binom{n-d}{k-d}det(X^TX)}
+            \end{equation}
+
+            This resultant $w_{S}$ is unbiased, which in turn allows the averaging of multiple results to reduce the variance.
+            However, there is no high probability guarantee of a $(1+\epsilon)L(w^\ast)$ bound on $L(w_{S})$ for this procedure
+            (as shown in \hyperref[sec:no_mult_bound_vol]{the following section}).
+
+    \section{No Multiplicative Bound For Standard Volume Sampling}
+    \label{sec:no_mult_bound_vol}
+    Placeholder text
+
 \end{document}

### 2019-02-26 summary/paper-summary.tex
updated volume sampling introduction

@@
             This resultant $w_{S}$ is unbiased, which in turn allows the averaging of multiple results to reduce the variance.
+            This procedure can also be implemented efficiently using the \textit{"Reverse Iterative Sampling"} and \textit{"Fast Regularized
+            Volume Sampling"} {\color{blue}\href{https://arxiv.org/abs/1710.05110}{\underline{algorithms}}}\footnote{\label{reg_vol_samp}"Subsampling
+            for Ridge Regression via Regularized Volume Sampling"}.
             However, there is no high probability guarantee of a $(1+\epsilon)L(w^\ast)$ bound on $L(w_{S})$ for this procedure
             (as shown in \hyperref[sec:no_mult_bound_vol]{the following section}).
+
+        \subsection{Leveraged Volume Sampling}
 

### 2019-02-27 run.sh
run.sh now shows debug prints by default

@@ -136,6 +136,8 @@ cache_file="${DEFAULT_DATASETS_CACHE}"
 graphs_directory="${DEFAULT_GRAPHS_DIRECTORY}"
 cpu_count=$(grep -cP '^processor' /proc/cpuinfo)
 
+export TRACE_INFO="1"
+
 while getopts ":hxtfn:o:c:d:s:" option
 do
     case "${option}" in
@@ -147,7 +149,7 @@ do
             setup_x_flag="-x"
             ;;
 
-        t)  export TRACE_INFO="1"
+        t)  export TRACE_INFO=""
             ;;
 
         f)  FORCE_FLATPAK="1"

### 2019-03-01 summary/paper-summary.tex
added volume sampling lower bounds summary

@@
             Assume the rows of such a sub-sampled problem to be $S$, the sub-sampled matrix and labels vector to be $X_{S}$
-            and $y_{S}$ respectively and the sub-sampled solution to be $w_{S}=X_{S}^+y_{S}$.
+            and $y_{S}$ respectively and the sub-sampled solution to be $w_{S}^\ast=X_{S}^+y_{S}$.
 
-            We'd like an effective algorithm to choose $S$, based on the features alone, whose corresponding solution $w_{S}$
+            We'd like an effective algorithm to choose $S$, based on the features alone, whose corresponding solution $w_{S}^\ast$
             will hold the following properties with regard to the optimal solution $w^\ast$:
             \begin{itemize}
-                \item $w_{S}$ is unbiased, i.e.: $E[w_{S}]=w^\ast$.
-                \item The error for $w_{S}$ can be bounded by $L(w_{S}) < (1+\epsilon)L(w^\ast)$ for an $\epsilon$ of choice.
+                \item $w_{S}^\ast$ is unbiased, i.e.: $E[w_{S}^\ast]=w^\ast$.
+                \item The error for $w_{S}^\ast$ can be bounded by $L(w_{S}^\ast) < (1+\epsilon)L(w^\ast)$ for an $\epsilon$ of choice.
                 \item Can be executed efficiently.
@@
             sampling for linear regression" P.21} that it is sufficient to choose
-            $O(d\log{d}+\frac{d}{\epsilon})$ to obtain $L(w_{S}) < (1+\epsilon)L(w^\ast)$ with high probability.
+            $\mathcal{O}(d\log{d}+\frac{d}{\epsilon})$ to obtain $L(w_{S}^\ast) < (1+\epsilon)L(w^\ast)$ with high probability.
             All in all this is a pretty efficient and fast algorithm.
-            However, this multiplicative bound doesn't hold in expectation and the approximate solution $w_{S}$ is biased (a
+            However, this multiplicative bound doesn't hold in expectation and the approximate solution $w_{S}^\ast$ is biased (a
             trait of i.i.d sampling algorithms). This means that you can't control for the error by repeating the process
@@
 
-            This resultant $w_{S}$ is unbiased, which in turn allows the averaging of multiple results to reduce the variance.
+            This resultant $w_{S}^\ast$ is unbiased, which in turn allows the averaging of multiple results to reduce the variance.
             This procedure can also be implemented efficiently using the \textit{"Reverse Iterative Sampling"} and \textit{"Fast Regularized
@@
             for Ridge Regression via Regularized Volume Sampling"}.
-            However, there is no high probability guarantee of a $(1+\epsilon)L(w^\ast)$ bound on $L(w_{S})$ for this procedure
+            However, there is no high probability guarantee of a $(1+\epsilon)L(w^\ast)$ bound on $L(w_{S}^\ast)$ for this procedure
             (as shown in \hyperref[sec:no_mult_bound_vol]{the following section}).
@@
         \subsection{Leveraged Volume Sampling}
+            \textit{Leveraged Volume Sampling} is this paper's contribution and it defines an unbiased sampling procedure with high probability
+            guarantees of multiplicative bounds for a relatively small $k$.
+            The procedure suggests sampling $k$ lines from $X$ with repetition in the following proportion:
+
+            \begin{equation}
+                P(\pi\in{[n]^k})\propto{det(\sum_{i=1}^{k}\frac{d}{l_{\pi_{i}}}x_{\pi_{i}}x_{\pi_{i}}^T)\prod_{i=1}^{k}\frac{l_{\pi_{i}}}{d}}, \text{where } l_{i}=x_{i}^T(X^TX)^{-1}x_{i}
+            \end{equation}
 
@@
     \label{sec:no_mult_bound_vol}
-    Placeholder text
+    First, the paper first demonstrates that for the following $n x d$ regression problem:
+    \begin{equation*}
+        X=
+            \begin{bmatrix}
+                I_{dxd} \\
+                \gamma I_{dxd} \\
+                \vdots \\
+                \gamma I_{dxd}
+            \end{bmatrix}
+        , y=
+            \begin{bmatrix}
+                1_{dxd} \\
+                0_{d} \\
+                \vdots \\
+                0_{d}
+            \end{bmatrix}
+        , \text{where } \gamma > 0
+    \end{equation*}
+
+    The paper proves that if $S$ of size $k$ is picked via ordinary \textit{Volume Sampling} then the following holds:
+    \begin{equation}
+        \lim\limits_{\gamma \to 0} \frac{E(L(w_{S}^\ast))}{L(w^\ast)} \geq 1+\frac{n-k}{n-d}
+    \end{equation}
+    \begin{equation}
+        \exists \gamma > 0: \forall k \leq \frac{n}{2}: P(L(w_{S}^\ast) \geq (1+\frac{1}{2})L(w^\ast)) > \frac{1}{4}
+    \end{equation}
+
+    This means that for \textit{Volume Sampling}:
+    \begin{itemize}
+        \item There can be no unconditional guarantee on the unbiasedness of the loss $L(w_{S}^\ast)$.
+        \item There can be no unconditional upper bound guarantee with high probability on the required $k$ which is less than $\mathcal{O}(n)$.
+    \end{itemize}
+
+    \textit{Leveraged Volume Sampling} does away with these limitations.
 

### 2019-03-01 summary/paper-summary.tex
wip - rescaled vol sampling summary

@@
             \begin{equation}
-                P(\pi\in{[n]^k})\propto{det(\sum_{i=1}^{k}\frac{d}{l_{\pi_{i}}}x_{\pi_{i}}x_{\pi_{i}}^T)\prod_{i=1}^{k}\frac{l_{\pi_{i}}}{d}}, \text{where } l_{i}=x_{i}^T(X^TX)^{-1}x_{i}
+                P(\pi\in{[n]^k}) \propto det(\sum_{i=1}^{k}\frac{d}{l_{\pi_{i}}}x_{\pi_{i}}x_{\pi_{i}}^T)\prod_{i=1}^{k}\frac{l_{\pi_{i}}}{d}, \text{where } l_{i}=x_{i}^T(X^TX)^{-1}x_{i}
             \end{equation}
@@
 
+    \section{Rescaled Volume Sampling Analysis}
+        \textit{Leveraged Volume Sampling} is a specific case of \textit{Rescaled Volume Sampling}.
+        Assume:
+        \begin{equation*}
+            \forall 1 \leq i \leq n: 0 \le q_{i} \leq 1, \text{ and } \sum_{i=1}^{n}q_{i} = 1
+        \end{equation*}
+        Then $(q_{i})_{i=1}^n$ defines a distribution over the lines of $X$.
+        \textit{Rescaled Volume Sampling} suggests sampling $k$ rows with repetition from $X$ with the following proportion:
+        \begin{equation}
+            P(\pi \in [n]^k) \propto det(\sum_{i=1}^{k}\frac{1}{q_{\pi_{i}}}x_{\pi_{i}}x_{\pi_{i}}^T)\prod_{i=1}^{k}q_{\pi_{i}} = det(X^{T}Q_{\pi}X)\prod_{i=1}^{k}q_{\pi_{i}},
+            \text{ where } Q=\sum_{i=1}^{k}\frac{1}{q_{\pi_{i}}}e_{\pi_{i}}e_{\pi_{i}}^T
+        \end{equation}
+        \begin{equation}
+            w_{\pi}^\ast=(Q_{\pi}^{1/2}X)^{+}Q_{\pi}^{1/2}y
+        \end{equation}
+
+        By application of the following theorems the paper proves that $w_{\pi}^\ast$ is unbiased:
+        \begin{itemize}
+            \item The Cauchy-Binet theorem, decomposing $det(XQ_{\pi}X)$ as the sum of determinants of square sub-matrices of $X$, to calculate $\sum_{\pi \in [n]^k}det(X^{T}Q_{\pi}X)\prod_{i=1}^{k}q_{\pi_{i}}$.
+            \item $E((I_{S}X)^+) = X^+$ for ordinary volume sampling to decompose both $w_{\pi}^\ast$ and $w^\ast$.
+        \end{itemize}
+
 \end{document}

### 2019-03-02 summary/paper-summary.tex
wip - lev. vol sampling section

@@
 \usepackage{amsmath}
+\usepackage{mathtools}
 
@@
 
+        Furthermore, using a bit of algebra, the paper shows that for \textit{Rescaled Volume Sampling}:
+        \begin{equation}
+            E(Q_{\pi}) = (k-d)I + diag(\frac{l_{1}}{q_{1}}, \hdots, \frac{l_{n}}{q_{n}})
+        \end{equation}
+
+    \section{Leveraged Volume Sampling}
+        \textit{Leveraged Volume Sampling} is \textit{Rescaled Volume Sampling} where $q_{i}=\frac{l_{i}}{d}$ (note that $\sum_{i=1}^{n}l_{i}=d$).
+        This procedure is unbiased (as it is a case of \textit{Rescaled Volume Sampling}) and is subject to favorable tail bounds (more on that in the next section).
+        You can poll such a subset with an algorithm named \textit{Determinantal Rejection Sampling} which selects $k$ lines from $X$ that boils down to the following steps:
+        \begin{itemize}
+            \item Set $s \coloneqq max(k, 4d^{2})$.
+            \item Repeat until acceptance selection of lines $\pi$ of size $s$ i.i.d from $1..n$ and accept with probability $\frac{det(\frac{1}{s}X^{T}Q_{\pi}X)}{det(X^{T}X)}$.
+            \item Use \textit{Volume Sampling} to reduce the selected $s$ rows to $k$ rows.
+        \end{itemize}
+
+        The proof that this procedure indeed yields a \textit{Leveraged Volume Sampling} distributed set is largely due to the fact that when you
+        select a \textit{Leveraged Volume Sampling} distributed set of size $s$ and then apply \textit{Reverse Iterative Volume Sampling} to sub-select
+        a $k$ sized subset then the \textit{Leveraged Volume Sampling} distribution is recursively maintained at each step.
+
 \end{document}

### 2019-03-02 summary/paper-summary.tex
finished the lev. vol sampling section

@@
 
+        This algorithm runs, with probability at least $1-\delta$, in time $\mathcal{O}((d^2+k^2)d^2ln(\frac{1}{\delta}))$.
+        This can be seen by bounding from below the acceptance probability in the repetition with $\frac{3}{4}$ and noting that determinant
+        calculation takes at most as much as decomposing a matrix via SVD which is $\mathcal{O}(sd^2)$.
+        Finally, note that $s=\mathcal{O}(k+d^2)$.
+
 \end{document}

### 2019-03-02 summary/paper-summary.tex
added a tail bounds summary

@@
 
+    \section{Tail Bounds For Leveraged Volume Sampling}
+        To analyze tail bounds the paper shifts to assesing the problem where instead of $X=(x_{i}^T)_{i=1}^{n}$ you have an orthogonal $U=(u_{i}^T)_{i=1}^{n}, U^TU=I$ that spans the range.
+        This maintains correctness due to the following properties:
+        \begin{itemize}
+            \item The leverage scores are equal: $\forall i: x_{i}^T(X^TX)^{-1}x_{i}=u_{i}^T(U^TU)^{-1}u_{i}$
+            \item The \textit{Volume Sampling} volumes are proportional: $\forall S: det(X_{S}^TX_{S}) \propto det(U_{S}^TU_{S})$
+            \item The \textit{Leveraged Volume Sampling} volumes are proportional: $\forall \pi: det(X^TQ_{\pi}X) \propto det(U^TQ_{\pi}U)$
+        \end{itemize}
+
+        The following notation changes hold:
+        \begin{itemize}
+            \item $v^\ast=argmin||Uv - y||$
+            \item $v_{\pi}^\ast=argmin||U_{\pi}v - y_{\pi}||$
+            \item $L(v) = ||Uv - y||, L(v^\ast) = L(w^\ast), L(v_{\pi}^\ast) = L(w_{\pi}^\ast)$
+            \item $r \coloneqq Uv^\ast - y, U^Tr = 0$
+        \end{itemize}
+
+        Obtaining a multiplicative bound on $L(v_{\pi}^\ast)$ is done by reducing the problem to:
+        \begin{equation}
+            ||v_{\pi}^\ast - v^\ast|| \leq ||(U^TQ_{\pi}U)^{-1}||||U^TQ_{\pi}r||
+        \end{equation}
+        And bounding these two factors.
+        By the following two traits we can get a multiplicative bound of $L(w_{\pi}^\ast) < (1+\epsilon)L(w^\ast)$,
+        with high probability, when $k=\Omega(d\log \frac{d}{\delta} + \frac{d}{\epsilon\delta})$:
+        \begin{itemize}
+            \item if $k \geq \frac{2d}{\epsilon}$ then $E(||\frac{1}{k}U^TQ_{\pi}r - U^Tr||^2) \leq \epsilon||r||^2$
+            \item There exists a $C$ s.t. if $k \geq Cd\log(\frac{d}{\delta})$ then $P(\lambda_{min}(\frac{1}{k}U^TQ_{\pi}U) \leq \frac{1}{8}) \leq \delta$
+        \end{itemize}
+
+        The first property, when applied to $r=U^Tv^\ast - y$, allows us to bound $U^TQ_{\pi}r$ and the second allows us to bound $||(U^TQ_{\pi}U)^{-1}|| < 8$
+        with high probability.
+
 \end{document}

### 2019-03-02 summary/paper-summary.tex
added results to the summary

@@
 \usepackage{mathtools}
+\usepackage{graphicx}
 
@@
 
-        The first property, when applied to $r=U^Tv^\ast - y$, allows us to bound $U^TQ_{\pi}r$ and the second allows us to bound $||(U^TQ_{\pi}U)^{-1}|| < 8$
-        with high probability.
+        The first property, when applied to $r=Uv^\ast - y$, allows us to bound $||U^TQ_{\pi}r||^2 \leq \epsilon\frac{k^2}{8^2}||r||^2$
+        and the second allows us to bound $||(U^TQ_{\pi}U)^{-1}|| < 8$, both with high probability, there by yielding the bound:
+        \begin{equation}
+            L(w_{\pi}^\ast) \leq (1+\epsilon)L(w^\ast)
+        \end{equation}
+
+    \newpage
+    \section{Numerical Results}
+        I've done my best to replicate the results in the paper.
+        The trend of diminishing error at relatively small number of rows is there but I couldn't replicate the smooth results the paper presented.
+
+        \begin{figure}[h]
+            \includegraphics[width=\linewidth]{results/abalone.png}
+            \caption{abalone dataset.}
+            \label{fig:dataset1}
+        \end{figure}
+        \begin{figure}[h]
+            \includegraphics[width=\linewidth]{results/bodyfat.png}
+            \caption{bodyfat dataset.}
+            \label{fig:dataset1}
+        \end{figure}
+        \begin{figure}[h]
+            \includegraphics[width=\linewidth]{results/cadata.png}
+            \caption{cadata dataset.}
+            \label{fig:dataset1}
+        \end{figure}
+        \begin{figure}
+            \includegraphics[width=\linewidth]{results/cpusmall.png}
+            \caption{cpusmall dataset.}
+            \label{fig:dataset1}
+        \end{figure}
+        \begin{figure}
+            \includegraphics[width=\linewidth]{results/housing.png}
+            \caption{housing dataset.}
+            \label{fig:dataset1}
+        \end{figure}
+        \begin{figure}
+            \includegraphics[width=\linewidth]{results/mg.png}
+            \caption{mg dataset.}
+            \label{fig:dataset1}
+        \end{figure}
 

### 2019-03-02 summary/paper-summary.tex
added a brief about the main idea behind the proof of theorem 8

@@
 
+        To prove the first property, the paper demonstrates:
+        \begin{equation}
+            E(||\frac{1}{k}U^TQ_{\pi}r - U^Tr||^2) = r^TE((\frac{1}{k}Q_{\pi}-I)UU^T(\frac{1}{k}Q_{\pi}-I))r \leq ||E((\frac{1}{k}Q_{\pi}-I)UU^T(\frac{1}{k}Q_{\pi}-I))||||r||^2
+        \end{equation}
+        And proceeds to bound $||E((\frac{1}{k}Q_{\pi}-I)UU^T(\frac{1}{k}Q_{\pi}-I))||$ via a clever application of a Hadamard Product inequality.
+
         The first property, when applied to $r=Uv^\ast - y$, allows us to bound $||U^TQ_{\pi}r||^2 \leq \epsilon\frac{k^2}{8^2}||r||^2$

### 2019-03-02 summary/paper-summary.tex
added a brief on the proof of theorem 9

@@
 
+        The proof of the second property is more complex but at it's core it builds on a lemma (proved in the paper) that states that if
+        you sub-sample a set $T$ fixed number of rows uniformly without replacement from $m$ rows sampled via \textit{Leveraged Volume Sampling}
+        then you can bound the following:
+        \begin{equation}
+            P(\lambda_{min}(\frac{1}{|T|}U^TQ_{\pi_{T}}U) \leq \frac{1}{4}) \leq \delta
+        \end{equation}
+        Then, they procede to generate such a uniform set $T$ and a standard volume-sampled $S$ of size $k$ out of $[m]$ where $T \subseteq S$.
+        With high probability $|S \setminus T|=\mathcal{O}(d\log(d/\delta))$ and this in turn allows us to generate the bound in the second property.
+        To prove this property the paper makes use of three lemmas where they use Tchernoff bounds on matrices as an important tool.
+
         The first property, when applied to $r=Uv^\ast - y$, allows us to bound $||U^TQ_{\pi}r||^2 \leq \epsilon\frac{k^2}{8^2}||r||^2$

### 2019-03-04 src/main.m
now plotting the average loss instead of the loss of the average

@@ -60,8 +60,8 @@ function [optimal_l2error, regression_time, ...
                    dataset_name, sample_sz, sampling_count, rows(X), columns(X),
                    sl2error, optimal_l2error, sl2error_avg, sl2error_std);
         lss_sl2errors = [lss_sl2errors sl2error];
-        lss_sl2errors_avg = [lss_sl2errors sl2error_avg];
-        lss_sl2errors_std = [lss_sl2errors sl2error_std];
+        lss_sl2errors_avg = [lss_sl2errors_avg sl2error_avg];
+        lss_sl2errors_std = [lss_sl2errors_std sl2error_std];
 
         info_trace("\t%s: VSS(s=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X));
         [sw, sXw, sl2error, sl2error_avg, sl2error_std] = volume_sampling(X, y, sample_sz, sampling_count);
@@ -69,8 +69,8 @@ function [optimal_l2error, regression_time, ...
                    dataset_name, sample_sz, sampling_count, rows(X), columns(X),
                    sl2error, optimal_l2error, sl2error_avg, sl2error_std);
         vss_sl2errors = [vss_sl2errors sl2error];
-        vss_sl2errors_avg = [vss_sl2errors sl2error_avg];
-        vss_sl2errors_std = [vss_sl2errors sl2error_std];
+        vss_sl2errors_avg = [vss_sl2errors_avg sl2error_avg];
+        vss_sl2errors_std = [vss_sl2errors_std sl2error_std];
 
         info_trace("\t%s: LVSS(k=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X));
         [sw, sXw, sl2error, sl2error_avg, sl2error_std] = leveraged_volume_sampling(X, y, sample_sz, sampling_count);
@@ -78,8 +78,8 @@ function [optimal_l2error, regression_time, ...
                    dataset_name, sample_sz, sampling_count, rows(X), columns(X),
                    sl2error, optimal_l2error, sl2error_avg, sl2error_std);
         lvss_sl2errors = [lvss_sl2errors sl2error];
-        lvss_sl2errors_avg = [lvss_sl2errors sl2error_avg];
-        lvss_sl2errors_std = [lvss_sl2errors sl2error_std];
[…]

### 2019-03-04 src/check-results.m
updated the output graphs and added an auxiliary script to validate graphs

@@ -1,0 +1,59 @@
+#!/usr/bin/octave -fq
+
+% this script loads the explicit results generated by main.m and calculates the average
+% error, the standard deviation and the errors themselves.
+
+% let Octave know this is not a function file (just write any word other than 'function').
+1;
+
+function process_solutions(X, y, solutions ,label)
+    % sort the solutions into buckets by the sample size
+    solutions_store = {};
+    sample_sizes = [];
+
+    for sample_size=sort(unique(solutions(1,:)))
+        solutions_store{sample_size} = [];
+        sample_sizes = [sample_sizes sample_size];
+    endfor
+
+    for i=1:columns(solutions)
+        solution = solutions(:, i);
+        solutions_store{solution(1,:)} = [solutions_store{solution(1,:)} solution(2:end,:)];
+    endfor
+
+    % now for each sample size calculate the losses and print their mean, std and errors
+    for sample_size=sample_sizes
+        solutions = solutions_store{sample_size};
+        l2errors = arrayfun(@(i) (norm(X*solutions(:,i) - y, 2)^2)/rows(X), 1:columns(solutions));
+        err_avg = mean(l2errors);
+        err_std = std(l2errors);
+
+        error_str = sprintf("%g ", l2errors);
+        printf("%s: %i samples stats: average=%g, std=%g, errors=[ %s]\n", label, sample_size, err_avg, err_std, error_str);
+    endfor
+
+endfunction
+
+function _main
+    % read command line arguments: the dataset file and the octave state
+    % file saved by the execution process
+    dataset_file = argv{1};
+    state_file = argv{2};
+
+    % load the state file so we can access it's stored variables
+    load(state_file);
+
+    % load the dataset
+    M = dlmread(dataset_file);
+    X = M(:, 2:end);
+    y = M(:, 1);
+
+    % now process the solutions
+    process_solutions(X, y, lss_solutions, "LSS");
+    process_solutions(X, y, vss_solutions, "VSS");
+    process_solutions(X, y, vlss_solutions, "VLSS");
+
+endfunction
+
+_main
+

### 2019-03-04 src/leverage_score_sampling.m

@@ -7,7 +7,7 @@
 %   * The averaged least-squares error (|sXw - y|^2).
 %   * The standard deviation of the least-squares error (|sXw - y|^2).
 
-function [sw, sXw, sl2error, sl2error_avg, sl2error_std]=leverage_score_sampling(X, y, k, times=1)
+function [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions]=leverage_score_sampling(X, y, k, times=1)
     n = rows(X);
     sw = zeros(columns(X), 1);
     sXw = zeros(rows(X), 1);
@@ -20,6 +20,8 @@ function [sw, sXw, sl2error, sl2error_avg, sl2error_std]=leverage_score_sampling
     leverage_score_sampler = LeverageScoreDistribution(X, y);
     info_trace("sampler initialized!")
 
+    solutions = [];
+
     for t=1:times
         info_trace("sampling iteration %i", t);
         % sample a subset of the rows via leverage score sampling
@@ -27,6 +29,7 @@ function [sw, sXw, sl2error, sl2error_avg, sl2error_std]=leverage_score_sampling
 
         % regress
         [_sw, _sXw, _sl2error, regression_time] = linear_regression(sX, sy);
+        solutions = [solutions _sw];
 
         % calculate the actual estimation (_sXw and _sl2error are irrelevant as they represent a sub-sampled problem)
         _sXw = X*_sw;

### 2019-03-04 src/leveraged_volume_sampling.m

@@ -9,7 +9,7 @@
 %   * The average l2 error for each of the `times` runs.
 %   * The standard deviation of the l2 error for each of the `times` runs.
 
-function [sw, sXw, sl2error, sl2error_avg, sl2error_std] = leveraged_volume_sampling(X, y, k, times)
+function [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = leveraged_volume_sampling(X, y, k, times)
     info_trace("initializing leveraged volume sampler for %ix%i matrix", rows(X), columns(X))
 
     n = rows(X);
@@ -23,6 +23,7 @@ function [sw, sXw, sl2error, sl2error_avg, sl2error_std] = leveraged_volume_samp
     sl2error_std = 0;
     total_time = 0;
     sl2errors = [];
+    solutions = [];
 
     for t = 1:times
         info_trace("LVSS iteration #%i", t);
@@ -41,6 +42,7 @@ function [sw, sXw, sl2error, sl2error_avg, sl2error_std] = leveraged_volume_samp
         info_trace("\tregressing...");
         _sw = linear_regression(_sX, _sy);
 
+        solutions = [solutions _sw];
         sw += _sw;
 
         _sl2error = (norm(X*_sw - y, 2)^2)/n;

### 2019-03-04 src/main.m

@@ -40,21 +40,24 @@ function [optimal_l2error, regression_time, ...
     lss_sl2errors = [];
     lss_sl2errors_avg = [];
     lss_sl2errors_std = [];
+    lss_solutions = [];
 
     vss_sl2errors = [];
     vss_sl2errors_avg = [];
     vss_sl2errors_std = [];
+    vss_solutions = [];
 
     lvss_sl2errors = [];
     lvss_sl2errors_avg = [];
     lvss_sl2errors_std = [];
+    lvss_solutions = [];
 
     optimal_error = optimal_l2error * ones(size(sample_sizes));
 
     info_trace ("proceeding to sample via LSS, VSS and LVSS for %i tries each", sampling_count)
     for sample_sz = sample_sizes
         info_trace("\t%s: LSS(k=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X))
-        [sw, sXw, sl2error, sl2error_avg, sl2error_std] = leverage_score_sampling(X, y, sample_sz, sampling_count);
+        [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = leverage_score_sampling(X, y, sample_sz, sampling_count);
 
         info_trace("%s: LSS(k=%i, times=%i, n=%i, d=%i): sl2error=%f, ol2error=%f, sl2error_avg=%f[%f]\n",
                    dataset_name, sample_sz, sampling_count, rows(X), columns(X),
@@ -62,24 +65,30 @@ function [optimal_l2error, regression_time, ...
         lss_sl2errors = [lss_sl2errors sl2error];
         lss_sl2errors_avg = [lss_sl2errors_avg sl2error_avg];
         lss_sl2errors_std = [lss_sl2errors_std sl2error_std];
+        solutions = [sample_sz*ones(1, columns(solutions)); solutions];
+        lss_solutions = [lss_solutions solutions];
 
         info_trace("\t%s: VSS(s=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X));
-        [sw, sXw, sl2error, sl2error_avg, sl2error_std] = volume_sampling(X, y, sample_sz, sampling_count);
+        [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = volume_sampling(X, y, sample_sz, sampling_count);
[…]

### 2019-03-04 src/volume_sampling.m

@@ -8,7 +8,7 @@
 %   * The average l2 error for each of the `times` runs.
 %   * The standard deviation of the l2 error for each of the `times` runs.
 
-function [sw, sXw, sl2error, sl2error_avg, sl2error_std] = volume_sampling(X, y, k, times)
+function [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = volume_sampling(X, y, k, times)
     info_trace("initializing volume sampler for %ix%i matrix", rows(X), columns(X))
 
     n = rows(X);
@@ -22,6 +22,7 @@ function [sw, sXw, sl2error, sl2error_avg, sl2error_std] = volume_sampling(X, y,
     sl2error_std = 0;
     total_time = 0;
     sl2errors = [];
+    solutions = [];
 
     for t = 1:times
         info_trace("VSS iteration #%i", t);
@@ -39,6 +40,7 @@ function [sw, sXw, sl2error, sl2error_avg, sl2error_std] = volume_sampling(X, y,
 
         info_trace("\tregressing...");
         _sw = linear_regression(sX, sy);
+        solutions = [solutions _sw];
 
         sw += _sw;
 

### 2019-03-04 src/check-results.m
fixed a typo bug in check-results.m

@@ -51,7 +51,7 @@ function _main
     % now process the solutions
     process_solutions(X, y, lss_solutions, "LSS");
     process_solutions(X, y, vss_solutions, "VSS");
-    process_solutions(X, y, vlss_solutions, "VLSS");
+    process_solutions(X, y, lvss_solutions, "LVSS");
 
 endfunction
 

### 2019-03-08 setup.sh
added feature squaring for certain datasets

@@ -10,6 +10,9 @@ set -eu
 readonly PROJECT_BASE=$(dirname $(readlink -f "${BASH_SOURCE[0]}"))
 readonly SCRIPT=$(basename "${BASH_SOURCE[0]}")
 
+readonly SOURCE_DIR="${PROJECT_BASE}/src"
+readonly SQUARE_FEATURES_SCRIPT="${SOURCE_DIR}/square-features.m"
+
 readonly DEFAULT_DATASETS_FILE="${PROJECT_BASE}/datasets-config"
 readonly DEFAULT_DATASETS_DIR="${PROJECT_BASE}/downloaded-datasets/"
 readonly DEFAULT_DATASETS_CACHE="${PROJECT_BASE}/datasets-cache"
@@ -127,6 +130,47 @@ function update_cached_file()
     write_cache_line "${_cache_file}" "${_dataset_url}" "${_output_file}" "${_cached_hash}" "${_sampling_count}"
 }
 
+# octave detection functions
+use_flatpak_octave=""
+function run_octave()
+{
+    if [ -z "${use_flatpak_octave}" ]
+    then
+        perror "Octave version was not detected"
+    elif [ "${use_flatpak_octave}" == "1" ]
+    then
+        flatpak run                         \
+            --branch=stable                 \
+            --arch=x86_64                   \
+            --command=/app/bin/octave-cli   \
+            --filesystem=host               \
+            org.octave.Octave "${@}"
+    else
+        octave-cli "${@}"
+    fi
+}
+
+function detect_octave()
+{
+    # check the path for octave
+    if [ -z "${FORCE_FLATPAK:-}" ] && which "octave-cli" &> /dev/null
+    then
+        use_flatpak_octave="0"
+        return 0;
+    fi
+
+    # the path doesn't hold a good enough octave.
+    # check flatpak.
+    if which flatpak &>/dev/null && flatpak list | grep -qiw Octave
+    then
+        use_flatpak_octave="1"
+        return 0;
+    fi
+
+    use_flatpak_octave=""
+    return 1
+}
+
 # download a dataset file
 function fetch_dataset()
 {
@@ -134,6 +178,7 @@ function fetch_dataset()
     local _dataset_url="${2}"
     local _output_directory="${3}"
     local _sampling_count="${4}"
+    local _square_features="${5}"
 
     local _output_file="${_output_directory}/"$(basename "${_dataset_url}" | awk -F/ '{print $NF}')
[…]

### 2019-03-08 src/square-features.m

@@ -1,0 +1,16 @@
+#!/usr/bin/octave -fq
+
+% this script loads a dataset, augments it with columns of the square of it's features and writes
+% the results back to the same dataset file.
+
+% let Octave know this is not a function file (just write any word other than 'function').
+1;
+
+function _main
+    dataset_file = argv{1};
+    M = dlmread(dataset_file);
+    dlmwrite(dataset_file, [M M(:,2:end).^2], " ");
+endfunction
+
+_main
+

### 2019-03-17 src/LeveragedVolumeSampler.m
added results for MSD

@@ -44,6 +44,10 @@ classdef LeveragedVolumeSampler < handle
             self._s = max(k, 4*(d^2));
             self._total_volume = det(X'*X);
 
+            if isnan(self._total_volume) || isinf(self._total_volume)
+                error("total volume is infinite!")
+            endif
+
             self._leverage_scores_distribution = LeverageScoreDistribution(X, y);
 
             if self._s >= self._n
@@ -62,7 +66,14 @@ classdef LeveragedVolumeSampler < handle
 
                 Q = diag(1./sqrt(scores));
                 _sX = Q*self._X(polled_rows, :);
-            until rand() < det((_sX'*_sX)/self._s)/self._total_volume
+
+                acceptance_probability = det((_sX'*_sX)/self._s)/self._total_volume;
+
+                if isnan(acceptance_probability) || isinf(acceptance_probability) || acceptance_probability < -0.0001
+                    info_trace("invalid acceptance probability: %g", acceptance_probability);
+                    error("invalid acceptance probability");
+                endif
+            until rand() < acceptance_probability
 
             % now do regular volume sampling on the polled rows
             if self._k < self._s

### 2019-03-17 src/info_trace.m

@@ -8,7 +8,7 @@ function info_trace(msg, varargin)
 
     if show_info_traces
         func_name = dbstack()(2).name;
-        msg = sprintf("info: %s [%s] %s: %s\n", prefix, strftime('%a %b %d %H:%M:%S %Y',gmtime(time())), func_name, msg);
+        msg = sprintf("info: %s [%s] %s: %s\n", prefix, strftime('%a %b %d %H:%M:%S %Y',localtime(time())), func_name, msg);
         fprintf(stderr(), msg, varargin{:});
     endif
 endfunction

### 2019-03-17 src/main.m

@@ -62,6 +62,17 @@ function [optimal_l2error, regression_time, ...
 
     info_trace ("proceeding to sample via LSS, VSS and LVSS for %i tries each", sampling_count)
     for sample_sz = sample_sizes
+        info_trace("\t%s: LVSS(k=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X));
+        [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = leveraged_volume_sampling(X, y, sample_sz, sampling_count);
+        info_trace("%s: LVSS(s=%i, times=%i, n=%i, d=%i): sl2error=%f, ol2error=%f, sl2error_avg=%f[%f]\n",
+                   dataset_name, sample_sz, sampling_count, rows(X), columns(X),
+                   sl2error, optimal_l2error, sl2error_avg, sl2error_std);
+        lvss_sl2errors = [lvss_sl2errors sl2error];
+        lvss_sl2errors_avg = [lvss_sl2errors_avg sl2error_avg];
+        lvss_sl2errors_std = [lvss_sl2errors_std sl2error_std];
+        solutions = [sample_sz*ones(1, columns(solutions)); solutions];
+        lvss_solutions = [lvss_solutions solutions];
+
         info_trace("\t%s: LSS(k=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X))
         [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = leverage_score_sampling(X, y, sample_sz, sampling_count);
 
@@ -84,17 +95,6 @@ function [optimal_l2error, regression_time, ...
         vss_sl2errors_std = [vss_sl2errors_std sl2error_std];
         solutions = [sample_sz*ones(1, columns(solutions)); solutions];
         vss_solutions = [vss_solutions solutions];
-
-        info_trace("\t%s: LVSS(k=%i, times=%i, n=%i, d=%i)", dataset_name, sample_sz, sampling_count, rows(X), columns(X));
-        [sw, sXw, sl2error, sl2error_avg, sl2error_std, solutions] = leveraged_volume_sampling(X, y, sample_sz, sampling_count);
-        info_trace("%s: LVSS(s=%i, times=%i, n=%i, d=%i): sl2error=%f, ol2error=%f, sl2error_avg=%f[%f]\n",
[…]

### 2019-03-17 src/main.m
updated the MSD graph to it's final form

@@ -27,7 +27,7 @@ function [optimal_l2error, regression_time, ...
 
     % normalize X and y
     info_trace("normalizing dataset...");
-    [X, y] = normalize_dataset(X, y);
+    [X, y, scaling_factor] = normalize_dataset(X, y);
 
     info_trace("dataset normalized");
 
@@ -97,6 +97,21 @@ function [optimal_l2error, regression_time, ...
         vss_solutions = [vss_solutions solutions];
     end
 
+    % now rescale everything with the scaling factor
+    optimal_error *= scaling_factor^2;
+
+    lvss_sl2errors *= scaling_factor^2;
+    lvss_sl2errors_avg *= scaling_factor^2;
+    lvss_sl2errors_std *= scaling_factor^2;
+
+    vss_sl2errors *= scaling_factor^2;
+    vss_sl2errors_avg *= scaling_factor^2;
+    vss_sl2errors_std *= scaling_factor^2;
+
+    lss_sl2errors *= scaling_factor^2;
+    lss_sl2errors_avg *= scaling_factor^2;
+    lss_sl2errors_std *= scaling_factor^2;
+
     info_trace("finished regressions - plotting results to %s", output_file);
 
     graph_handle = figure();
@@ -114,7 +129,7 @@ function [optimal_l2error, regression_time, ...
 
     print(graph_handle, output_file, "-dpng");
 
-    save(strcat(output_file, ".octave"), "sample_sizes", "lss_sl2errors", "vss_sl2errors", "lvss_sl2errors", "optimal_error", "lss_solutions", "vss_solutions", "lvss_solutions");
+    save(strcat(output_file, ".octave"), "scaling_factor", "sample_sizes", "lss_sl2errors", "vss_sl2errors", "lvss_sl2errors", "optimal_error", "lss_solutions", "vss_solutions", "lvss_solutions");
 
     info_trace("done plotting!");
 

### 2019-03-17 src/normalize_dataset.m

@@ -14,7 +14,7 @@
 % Note that we drop any eigenvalues that are diminished to 0 (which means either
 % they were 0 or there was some floating point error.
 
-function [nX, ny] = normalize_dataset(X, y)
+function [nX, ny, scaling_factor] = normalize_dataset(X, y)
     d = columns(X);
     Z = X'*X;
 
@@ -27,5 +27,6 @@ function [nX, ny] = normalize_dataset(X, y)
 
     nX = X/factor;
     ny = y/factor;
+    scaling_factor = factor;
 endfunction
 

### 2019-03-17 utils/refactor-MSD-graph.m

@@ -1,0 +1,74 @@
+% the MSD graph was generated after rescaling X and y - which cause the error to be greatly reduced.
+% this script loads the MSD saved state and generates a correct graph.
+
+1;
+
+function _main
+    fprintf(stderr(), "loading dataset\n");
+    M=dlmread("downloaded-datasets/YearPredictionMSD" );
+    fprintf(stderr(), "dataset loaded\n");
+    y=M(:,1);
+    X=M(:,2:end);
+
+    load("graphs/YearPredictionMSD.png.octave");
+
+    errorfunc = @(v) (norm(y - X*v, 2)^2)/rows(X);
+
+    lvss_errors = {};
+    vss_errors = {};
+    lss_errors = {};
+
+    for sz=sample_sizes
+        lvss_errors{sz} = [];
+        vss_errors{sz} = [];
+        lss_errors{sz} = [];
+    endfor
+    
+    for i=1:columns(lvss_solutions)
+        compound_solution = lvss_solutions(:,i);
+        sample_size = compound_solution(1,1);
+        solution = compound_solution(2:end,1);
+
+        fprintf(stderr(), "evaluating solution of size %i\n", sample_size);
+
+        lvss_errors{sample_size} = [lvss_errors{sample_size} errorfunc(solution)];
+
+        compound_solution = vss_solutions(:,i);
+        sample_size = compound_solution(1,1);
+        solution = compound_solution(2:end,1);
+        vss_errors{sample_size} = [vss_errors{sample_size} errorfunc(solution)];
+
+        compound_solution = lss_solutions(:,i);
+        sample_size = compound_solution(1,1);
+        solution = compound_solution(2:end,1);
+        lss_errors{sample_size} = [lss_errors{sample_size} errorfunc(solution)];
+    endfor
+
+    optimal_error = errorfunc(X\y)*ones(size(sample_sizes));
+
+    fprintf(stderr(), "calculating mean LVSS errors\n");
+    lvss_sl2errors_avg = arrayfun(@(sz) mean(lvss_errors{sz}), sample_sizes);
+    fprintf(stderr(), "calculating mean VSS errors\n");
+    vss_sl2errors_avg = arrayfun(@(sz) mean(vss_errors{sz}), sample_sizes);
+    fprintf(stderr(), "calculating mean LSS errors\n");
[…]
